G\"{o}del's Ontological Proof of God's Existence (Draft)

### 2013-09-17 Natural Deduction/GodProof-ND.tex
Add Annika's introduction and ND derivation of Modal Collapse

ToDo: Do also Scott's and Gödel's proofs.

@@

@@

\begin{theorem}
\label{T3}
Necessarily, God exists:
$$
\nec \ex x. G(x)
$$
\end{theorem}

\noindent
\textbf{Formal proof:}
Let the following derivation be $ \Pi$:

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec \ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec\nec \ex x. G(x)$}\RightLabel{$\imp_I^1$}
\UIC{$\nec \ex x. G(x)\imp\nec\nec \ex x. G(x)$} \RightLabel{propositional logic}
\UIC{$\neg\nec\nec \ex x. G(x)\imp\neg  \nec \ex x. G(x)$}\RightLabel{$\nec_I$}
\UIC{$\nec(\neg\nec\nec \ex x. G(x)\imp\neg  \nec \ex x. G(x))$}
\AXC{Axiom K} \dashedLine 
\BIC{$\nec\neg\nec\nec \ex x. G(x)\imp\nec\neg  \nec \ex x. G(x)$} \dottedLine \RightLabel{$\pos$}
 \UIC{$\nec\pos\neg\nec \ex x. G(x) \imp\nec\neg\nec \ex x. G(x) $}

\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec\pos\neg\nec \ex x. G(x)$}
\AXC{$ \Pi$} \dashedLine 
\UIC{$\nec\pos\neg\nec \ex x. G(x) \imp\nec\neg\nec \ex x. G(x) $}
 \BIC{$\nec\neg\nec \ex x. G(x)$}\RightLabel{$\nec_E$, axiom T}
\UIC{$\neg\nec \ex x. G(x)$} 
\AXC{$ $} \dashedLine  \RightLabel{Lemma \ref{L1}}
\UIC{$\neg\nec \ex x. G(x) \imp \neg\ex x. G(x)$}\RightLabel{$\imp_E$}
 \BIC{$ \neg\ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec \neg\ex x. G(x)$}  \dottedLine
\UIC{$\neg \pos\ex x. G(x)$}

\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $}\dashedLine
\UIC{$\neg \pos\ex x. G(x)$} 
\AXC{ Corollary \ref{C1}} \dashedLine 
\UIC{$ \pos\ex x. G(x)$} \RightLabel{$\imp_E$}
 \BIC{$ \bot$} \RightLabel{$\imp_I^1$}
 \UIC{$\neg\nec\pos\neg\nec \ex x. G(x)$}
\AXC{ Axiom B} \dashedLine 
\UIC{$ \neg\nec \ex x. G(x)\imp \nec\pos\neg\nec \ex x. G(x)$} \doubleLine
\UIC{$ \neg \nec\pos\neg\nec \ex x. G(x)\imp  \nec \ex x. G(x)$} \RightLabel{$\imp_E$}
 \BIC{$ \nec \ex x. G(x)$}
\end{prooftree}
\end{small}

\textbf{Better proof, that directly proves $\ex x. G(x)$:}
\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec\neg\nec \ex x. G(x)$}\RightLabel{$\nec_E$, axiom T}
\UIC{$\neg\nec \ex x. G(x)$} 
\AXC{ Lemma \ref{L1} } \dashedLine
\UIC{$\neg\nec \ex x. G(x) \imp \neg\ex x. G(x)$}\RightLabel{$\imp_E$}
 \BIC{$ \neg\ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec \neg\ex x. G(x)$}  \dottedLine
\UIC{$\neg \pos\ex x. G(x)$} 
\end{prooftree}
\end{small}

@@

\section{God exists}\label{GodsE}

@@

>>>>

ToDo: Do also Scott's and G\"{o}del's proofs.

The first ontological argument for the existence of God was written by St. Anselm (1033-1109). The argument considers a maximally concievable being. This being must exist, because if it did not have the property of existence, then we could concieve of a greater being that apart from the other properties also has the property of existence. The main critique of this argument is that we do not know whether the concept maximal concievable being in fact designates anything or if it is inconsistent, like a round square. The question is if the argument is sound.

As Bertrand Russell has pointed out the definition of maximal allows us to define properties, like having boots, which the maximal being then also must have. Kant, on the other hand argued against the ontological argument on the basis that existence is not an analytic property. This means that existence cannot be contained in the definition of a concept, because it is generally synthetic. All that we can say is that if God exists, then he necessarily exists.

St. Anselm's proof was processed by Descartes and Leibniz. Leibniz identified the critical point of the argument as establishing the possible existence of God. Leibniz gave an argument for that the properties of God, the perfections, are compatible. This implies that it is possible to have all perfections at once and therefore the existence of a maximal being with all these properties is possible.

G\"odel continued this argument by defining the properties that must hold for these perfections and giving axioms from which the existence of God can be derived. The acceptance of the correctness of the ontological argument by G\"odel's work boils down to the intuitive correctness of the axioms and definitions and the belief in the soundness of the deductive system. The formal argument of G\"odel is based on Leibniz proof, which in turn is based on Descartes proof. These proofs have two parts; a proof that if God's existence is possible, then it is necessary and a proof that God's existence is in fact possible. The deductive system used in the formal proof is a system for modal logic extended with second-order quantification (or possibly third-order quantification), because the axioms and definitions required for the concept of positive properties or perfections involve quantification over properties.

G\"odel treated positive properties as not just atomic properties, like Leibniz, but also consisting of collections of these properties, which by Leibniz argument are compatible or possible. (Fitting, p.139) G\"odel fomulated an axiom stating that the conjunction of any set of positive properties is positive. The property of God-likeness, which defines what a God is, must therefore be an infinite conjunction of all the positive properties. This axiom of the positiveness of conjunctions can be given a third order fomulation which is the reason for the claim that third-order quantification may be required. (Fitting, 2002, p. 148). However, this axiom is oly used in the proof to establish that the property of God-likeness is itself positive that led Scott to simply assume the positiveness of  God-likeness as an axiom instead.

Nevertheless, both axioms have the unfortunate property of being equivalent to the possibility of God's existence, given the other axioms of G\"odel. The possibility  God's existence in turn is equivalent to  God's necessary existence as well as  God's existence itself. Thus, we have an axiom which seems to be equivalent to the conclusion.

If one assumes extensionality, that two objects with the same properties are identical, then the definitions of G\"odel's proof give monotheism. Namely, two Gods would both be God-like and by the definition of god-like, they would have the same positive properties and be identical.

It may be so that for all formulas the formula implies its boxed version. That is all that is is necessarily and modality collapses. Fitting provides two solutions to the problem by modifying the axioms.

@@

\section{God exists}

@@

\begin{theorem}
\label{T4}
God exists:
$$
\ex x. G(x)
$$
\end{theorem}

\noindent
\textbf{Formal proof:}

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec\neg\nec \ex x. G(x)$}\RightLabel{$\nec_E$, axiom T}
\UIC{$\neg\nec \ex x. G(x)$} 
\AXC{ Lemma \ref{L1} } \dashedLine
\UIC{$\neg\nec \ex x. G(x) \imp \neg\ex x. G(x)$}\RightLabel{$\imp_E$}
 \BIC{$ \neg\ex x. G(x)$} \RightLabel{$\nec_I$}
\UIC{$\nec \neg\ex x. G(x)$}  \dottedLine
\UIC{$\neg \pos\ex x. G(x)$} 
\end{prooftree}
\end{small}

@@

We can also prove that god exists necessarily in our system by simply introducing box on a theorem by rule N in modal sytem K. So we get:

\begin{corollary}
\label{T3}
Necessarily, God exists:
$$
\nec \ex x. G(x)
$$
\end{corollary}

\section{God exists}\label{GodsE}
This section is superfluous.

@@

\section{Crisis in the calculus}

The axioms of the calculus imply the following collapse of modalities:

\begin{theorem}
\label{T5}
For all constant fomulas (without free variables), $A$,  we have:
$$
A \imp \nec A
$$
\end{theorem}

Note that in intuitionistic predicate logic we have $\all y.[B \imp C]  \biimp [\ex y.B \imp C]$ if $y$ is not free in $C$.

\begin{small}
\begin{prooftree}
\AXC{Theorem \ref{T2}}\dashedLine
\UIC{$\all y.[G(y) \imp \ess{G}{y}]$} \RightLabel{D2 }
\UIC{$\all y.[G(y) \imp G(y) \wedge \all \psi. (\psi(y) \imp \nec \all x. (G(x) \imp \psi(x)))]$} \RightLabel{Prop. logic }
\UIC{$\all y.[G(y) \imp \all \psi. (\psi(y) \imp \nec \all x. (G(x) \imp \psi(x)))]$} \RightLabel{Second order quantifier elimination and logical steps }
\UIC{$\all y.[G(y) \imp (A(y) \imp \nec \all x. (G(x) \imp A(x)))]$} \RightLabel{$A$ is constant}
\UIC{$\all y.[G(y) \imp (A \imp \nec \all x. (G(x) \imp A))]$} \RightLabel{intuitionistic predicate logic}
\UIC{$\ex y.G(y) \imp (A \imp \nec \all x. (G(x) \imp A))$}
\end{prooftree}
\end{small}
\begin{small}
\begin{prooftree}
\AXC{}\dashedLine
\UIC{$\ex y.G(y) \imp (A \imp \nec \all x. (G(x) \imp A))$}
\AXC{Theorem \ref{T4}}\dashedLine
\UIC{$\ex y.G(y)$} \RightLabel{$\imp_E$}
\BIC{$ A \imp \nec \all x. (G(x) \imp A)$} \RightLabel{intuitionistic predicate logic}
\UIC{$ A \imp \nec (\ex x. G(x) \imp A)$}
\AXC{$ $} \RightLabel{1}
\UIC{$A$} \RightLabel{$\imp_E$}
\BIC{$ \nec (\ex x. G(x) \imp A)$} \RightLabel{$\nec_E$}
\UIC{$ \ex x. G(x) \imp A$} \RightLabel{$\imp_E$}
\AXC{Theorem \ref{T4}}\dashedLine
\UIC{$\ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$A$}  \RightLabel{$\nec_I$}
\UIC{$ \nec A$}  \RightLabel{$\imp_I^1$}
   \UIC{$A\imp \nec A$}
\end{prooftree}
\end{small}

This collapse of the sytem is proved in (Fitting, 2002) and he provides two solutions by the modifications of the axioms, see chapter 11 section 8-11. The argument may require that one doesn't instantiate the second order quantifier with $A$ directly but with $\lambda x.A$ to make $A$ a property. But still we have $(\lambda x.A)(y)\equiv A$ because $A$ is constant.

The question arises what kind of second order quantification is allowed. The instantiations that are used in our proof are instantiated with predicate variables that later work as eigenvariables for the second order universal introduction. In the proof of corollary \ref{C1} we instantiate with $G$ and in the proof of lemma \ref{L1}.  we instantiate with $G$ and $E$.

In Sobel on G\"{o}del's ontological proof by Robert E. Koons this question is raised.

One can also discuss if a second order modal logic is sound and complete. With Henkin models, which restrict the second order quantification with comprehension, we have that normal second order logic is sound and complete. There is a completeness proof for a second order modal logic found in A completeness theorem in second order modal logic by Cocchiarella.
